Designing Types for R, Empirically

### 2020-05-15 paper/main.tex
Add 7.3

\input{corpus.tex}

@@

>>>>

\input{corpus.tex}
\input{macros/macros.tex}
\input{asserts.tex}

@@

There is a number of ways in R to check types of function parameters. The
default and the most common way\footnote{\CranStopifnotRatio of all runtime
  checks in the whole of CRAN} is to use \code{stopifnot} function from the R
base package. It takes a number of R expressions which all should evaluate to
true otherwise a runtime exception is thrown with a message quoting the
corresponding failed expression. For example, the following code checks whether
a given parameter \code{x} is a scalar string:
%
\begin{lstlisting}
stopifnot(is.character(x), length(x) == 1L, !is.na(x))
\end{lstlisting}

Next to \code{stopifnot}, there are 4 packages in CRAN\footnote{Packages are
  available on CRAN website: \url{https://cran.r-project.org/web/packages/}}.
that focus on runtime assertions: \emph{assertive}, \emph{ensurer},
\emph{assertr} and \emph{assertthat} \emph{assertive} and \emph{ensurer} have
not been updated 2016 and 2015 respectively. \emph{assertr} is used by only
\AssertrRevdeps other packages and in the current version focuses on checking
properties of data frames. Only the \emph{assertthat} is maintained and used
(with \AssertthatRevdeps reverse dependencies). The advantage of
\emph{assertthat} over the R's default is that it provides a much better error
message.

One way to asses the usefulness of our type checking system is to find out how
many of the existing type checking constrains could be replaced by \contractr.
To measure this, we have extracted all calls to \code{stopifnot} and
\code{assertthat} assertions and checked which one of them could be either
completely replaced by \contractr or at least partially simplified by removing a
portion of an assertion expression. This is useful, because a common pattern is
that the first part of the parameter assertion checks its type while the rest
checks its value. In the example above, the whole expression could be replaced
by \code{character} type.

Out of the \CorpusLoadable packages, \CorpusAssertsInPackages use runtime
assertions. Together there are \CorpusAsserts asserts in
\CorpusAssertsInFunctions functions. Among these, \contractr can replace
\CorpusTypedAsserts (\CorpusTypedAssertsRatio) assertion calls across
\CorpusTypedAssertsPackages packages and \CorpusTypedAssertsFunctions functions.
Furthermore, additional \CorpusPartiallyTypedAsserts
(\CorpusPartiallyTypedAssertsRatio) asserts in packages
\CorpusPartiallyTypedAssertsPackages packages and
\CorpusPartiallyTypedAssertsFunctions functions could have been simplified.

Checking the type of function parameters is not something that is seen often in
the R code. In the whole of CRAN, there are only \CranAssertsRnd asserts in
\CranAssertsInFunctionsRnd functions define in \CranAssertsInPackagesRnd
packages. One might speculate that one of the reasons is that the current way is
not convenient as it is rather verbose. Our system can infer type annotations
for existing functions automatically. This can remove or partially remove over
\CorpusPartiallyTypedAssertsRatio of existing assertions.
